---
title: '**Visualizing Household Data**'
author: "Xiang Wang"
output: html_document
---

#### Data Description
Measurements of electric power consumption in one household with a one-minute sampling rate over a period of almost 4 years. Different electrical quantities and some sub-metering values are available. The following descriptions of the 9 variables in the dataset are taken from the UCI web site:

* Date: Date in format dd/mm/yyyy
* Time: time in format hh:mm:ss
* Global_active_power: household global minute - averaged active power (in kilowatt)
* Global_reactive_power: household global minute - averaged reactive power (in kilowatt)
* Voltage: minute-averaged voltage (in volt)
* Global_intensity: household global minute-averaged current intensity (in ampere)
* Sub_metering_1: energy sub-metering No. 1 (in watt-hour of active energy). It corresponds to the kitchen, containing mainly a dishwasher, an oven and a microwave (hot plates are not electric but gas powered).
* Sub_metering_2: energy sub-metering No. 2 (in watt-hour of active energy). It corresponds to the laundry room, containing a washing-machine, a tumble-drier, a refrigerator and a light.
* Sub_metering_3: energy sub-metering No. 3 (in watt-hour of active energy). It corresponds to an electric water-heater and an air-conditioner.

#### Data Preparation

```{r}
# read a zip data file without unzipping it
data <- read.table(unz("household_power_consumption.zip", "household_power_consumption.txt"), header=T, sep=';') 
str(data) # check data structures and found most variables are not numeric
# read again to change the first two variables to character variables; variable 3 to variable 8 need to be changed from factors to numeric; stringsAsFactors=F tells R to keep character variables as they are rather than convert to factors
data <- read.table(unz("household_power_consumption.zip", "household_power_consumption.txt"), header=T, sep=';', nrows=2075259, na.strings="?", check.names=T, stringsAsFactors=F, comment.char="", quote='\"') 
str(data) # now all variables are appropriately attributed
head(data)[1:3,]

# suppose we are ONLY interested in the data from two days (2007-02-01 and 2007-02-02)
# first convert the Date and Time variables from format 'd/m/y' to 'y/m/d'
data$Date <- as.Date(data$Date, format="%d/%m/%Y")
## subset the data (only from the dates 2007-02-01 and 2007-02-02) 
data1 <- subset(data, subset=(Date >= "2007-02-01" & Date <= "2007-02-02"))
```

#### Visualizing Data
```{r}
# Plot 1 (Frequency of Global Active Power)
hist(data1$Global_active_power, main="Global Active Power", xlab="Global Active Power (kilowatts)", ylab="Frequency", col="Grey")

# Plot 2 (Time Series Graph of Global Active Power)
# get a continuos variable for a line plot
datetime <- paste(as.Date(data1$Date), data1$Time)
data1$Datetime <- as.POSIXct(datetime)
plot(data1$Global_active_power~data1$Datetime, type="l", ylab="Global Active Power (kilowatts)", xlab="")

# Plot 3 
with(data1, {
  plot(Sub_metering_1~Datetime, type="l", ylab="Energy sub metering", xlab="")
  lines(Sub_metering_2~Datetime,col='Red')
  lines(Sub_metering_3~Datetime,col='Blue')
})
legend("topright", pch=1, col=c("black", "red", "blue"), lty=1, lwd=2, 
       legend=c("Sub_metering_1", "Sub_metering_2", "Sub_metering_3"))

# Plot 4
par(mfrow=c(2,2), mar=c(4,4,2,1), oma=c(0,0,2,0))
with(data1, {
  plot(Global_active_power~Datetime, type="l", ylab="Global Active Power (kilowatts)", xlab="")
  plot(Voltage~Datetime, type="l", ylab="Voltage (volt)", xlab="")
  plot(Sub_metering_1~Datetime, type="l", ylab="Global Active Power (kilowatts)", xlab="")
  lines(Sub_metering_2~Datetime,col='Red')
  lines(Sub_metering_3~Datetime,col='Blue')
  legend("topright", col=c("black", "red", "blue"), lty=1, lwd=2, bty="n", legend=c("Sub_metering_1", "Sub_metering_2", "Sub_metering_3"))
  plot(Global_reactive_power~Datetime, type="l", ylab="Global Rective Power (kilowatts)",xlab="")
})
```
